Stochastic Variance-Reduced Policy Gradient

### 2018-02-09 svrpg.tex
Updated

The main limitation of plain policy gradient is the high variance of these estimators.
The na\"ive approach of increasing the batch size is not an option in RL due to the high cost of collecting samples, \ie by interacting with the environment.
For this reason, literature has focused on the introduction of baselines (\ie functions $b : \mathcal{S} \times \mathcal{A} \to \realspace$) aiming to reduce the variance~\citep[\eg][]{Peters2008reinf,Thomas2017actionbaseline,wu2018variance}.
There has been a surge of recent interest in variance reduction techniques for gradient optimization in supervised learning.
Although these techniques have been mainly derived for finite-sum problems, we will show in Section~\ref{sec:alg} how they can be used in RL.
The next section has the aim to describe variance reduction techniques for finite-sum problems. In particular, we will present the SVRG algorithm that is at the core of this work.

@@

\section{SVRG in Reinforcement Learning}\label{sec:alg}
In online RL problems, the usual approach is to tune the batch size of SG to find the optimal trade-off between variance and speed.
Recall that, compared to SL, the samples are not fixed in advance but we need to collect them at each policy change.
Since this operation may be costly, we would like to minimize the number of interactions with the environment.
For these reasons, we would like to apply SVRG to RL problems in order to limit the variance introduced by sampling trajectories, which would ultimately lead to faster convergence.
However, a direct application of SVRG to RL is not possible due to the following issues:
% In on-line policy gradient problems, SG is not really a choice, but is dictated by the necessity of interacting with an unknown environment.\todopir{I do not understand the meaning of this sentence. I would say that SG is not a choice by referring to papers that uses batches}
% We would like to apply SVRG to the policy gradient framework in order to limit the variance introduced by sampling trajectories, which would ultimately lead to faster convergence. In this scenario, the function $f(\vtheta)$ to optimize is $J(\vtheta)$ and the (implicit) dataset $\mathcal{D}$ is the set of all possible trajectories $\mathcal{T}$. Compared to the typical finite sum optimization scenario, we have the following additional challenges:
\begin{description}
        \item[Non-concavity:] the objective function $J(\vtheta)$ is typically non-concave.
        \item[Infinite dataset:] the RL optimization cannot be expressed as a finite-sum problem. The objective function is an expected value over the trajectory density $p_{\vtheta}(\tau)$ of the total discounted reward, for which we would need an infinite dataset.
%    \item[Approximation:] the dataset $\mathcal{T}$ can be infinite, but we can only sample a finite number of trajectories;
        \item[Non-stationarity:] the distribution of the samples changes over time. In particular, the value of the policy parameter $\vtheta$ influences the sampling process.
\end{description}
To deal with non-concavity, we require $J(\vtheta)$ to be $L$-smooth, which is a reasonable assumption for common policy classes such as Gaussian\footnote{see Appendix~\ref{app:gauss} for more details on the Gaussian policy case.} and softmax~\citep[\eg][]{Furmston2012unifying,pirotta2015lipschitz}.
Because of the infinite dataset, we can only rely on an estimate of the full gradient.
\citet{harikandeh2015stopwasting} analysed this scenario under the assumptions of $z$ being concave, showing that SVRG is robust to an inexact computation of the full gradient. In particular, it is still possible to recover the original convergence rate if the error decreases at an appropriate rate. In Section \ref{sec:conv} we will show how the estimation accuracy impacts on the convergence results with a non-concave objective.
Finally, the non-stationarity of the optimization problem introduces a bias into the SVRG estimator in Eq.~\eqref{E:svrg.gradient.correction}.
To overcome this limitation we employ importance weighting~\citep[\eg][]{rubinstein1981simulation,precup2000eligibility} to correct the distribution shift.
%Finally, we employ importance weighting~\citep[\eg]{rubinstein1981simulation,precup2000eligibility} to guarantee proper sampling of trajectories.

@@

but the snapshot gradient is not exact and the gradient estimate used between snapshots is corrected using importance weighting:

@@

There are been a surge of work investigating variance reduction techniques for policy gradient methods.
% The most widespread technique of reducing the variance preserving the unbiasedness of the gradient is through baselines~\citep[\eg][]{williams1992simple,weaver2001optimal,peters2008reinforcement,Thomas2017actionbaseline,wu2018variance}.
The specific structure of the policy gradient allows to incorporated a baseline (\ie a function $b :\mathcal{S}\times\mathcal{A} \to \realspace$) without affect the unbiasedness of the gradient~\citep[\eg][]{williams1992simple,weaver2001optimal,peters2008reinforcement,Thomas2017actionbaseline,wu2018variance}.
Although the baseline can be arbitrarily selected, literature often refer to the optimal baseline as the one minimizing the variance of the estimate.
Despite these approaches and SVRG has the same goal, they are substantially different.
In particular, the proposed SVRPG does not make explicit use of the structure of the policy gradient framework.
This suggests that would be possible to integrate an ad-hoc SVRPG baseline in order to further reduce the variance of the estimate.
But since this paper is about the applicability of SVRG technique to RL we consider this topic as future work.
Additionally, the experiments show the advantage of using simple SVRPG against G(PO)MPD with baseline (see experiment of \textcolor{blue}{XXX} in Section~\ref{sec:exp}).

@@

For G(PO)MDP, we do not have a compact expression for $g$, but since it is derived from REINFORCE by neglecting some terms of the kind $\nabla \log \pi_{\vtheta}{a_t | s_t}\gamma^t r_t$ \cite{baxter2001infinite,peters2008reinforcement}, the above bound still holds.

>>>>

The main limitation of plain policy gradient is the high variance of these estimators.
The na\"ive approach of increasing the batch size is not an option in RL due to the high cost of collecting samples, \ie by interacting with the environment.
For this reason, literature has focused on the introduction of baselines (\ie functions $b : \mathcal{S} \times \mathcal{A} \to \realspace$) aiming to reduce the variance~\citep[\eg][]{williams1992simple,Peters2008reinf,Thomas2017actionbaseline,wu2018variance}.
{\color{red}
This baselines are usually designed in order to minimize the variance of the gradient estimate but even them need to be estimated from data, partially reducing they effectiveness.
On the other hand, there has been a surge of recent interest in variance reduction techniques for gradient optimization in supervised learning.
Although these techniques have been mainly derived for finite-sum problems, we will show in Section~\ref{sec:alg} how they can be used in RL.
In particular, we will shown that the proposed SVRPG algorithm can take the best of both worlds (\ie SL and RL) since it can be plugged over a policy gradient estimate using baselines.
The next section has the aim to describe variance reduction techniques for finite-sum problems. In particular, we will present the SVRG algorithm that is at the core of this work.
}

@@

\section{SVRG in Reinforcement Learning}\label{sec:alg}
In online RL problems, the usual approach is to tune the batch size of SG to find the optimal trade-off between variance and speed.
Recall that, compared to SL, the samples are not fixed in advance but we need to collect them at each policy change.
Since this operation may be costly, we would like to minimize the number of interactions with the environment.
For these reasons, we would like to apply SVRG to RL problems in order to limit the variance introduced by sampling trajectories, which would ultimately lead to faster convergence.
However, a direct application of SVRG to RL is not possible due to the following issues:
% In on-line policy gradient problems, SG is not really a choice, but is dictated by the necessity of interacting with an unknown environment.\todopir{I do not understand the meaning of this sentence. I would say that SG is not a choice by referring to papers that uses batches}
% We would like to apply SVRG to the policy gradient framework in order to limit the variance introduced by sampling trajectories, which would ultimately lead to faster convergence. In this scenario, the function $f(\vtheta)$ to optimize is $J(\vtheta)$ and the (implicit) dataset $\mathcal{D}$ is the set of all possible trajectories $\mathcal{T}$. Compared to the typical finite sum optimization scenario, we have the following additional challenges:
\begin{description}
        \item[Non-concavity:] the objective function $J(\vtheta)$ is typically non-concave.
        \item[Infinite dataset:] the RL optimization cannot be expressed as a finite-sum problem. The objective function is an expected value over the trajectory density $p_{\vtheta}(\tau)$ of the total discounted reward, for which we would need an infinite dataset.
%    \item[Approximation:] the dataset $\mathcal{T}$ can be infinite, but we can only sample a finite number of trajectories;
        \item[Non-stationarity:] the distribution of the samples changes over time. In particular, the value of the policy parameter $\vtheta$ influences the sampling process.
\end{description}
To deal with non-concavity, we require $J(\vtheta)$ to be $L$-smooth, which is a reasonable assumption for common policy classes such as Gaussian\footnote{See Appendix~\ref{app:gauss} for more details on the Gaussian policy case.} and softmax~\citep[\eg][]{Furmston2012unifying,pirotta2015lipschitz}.
Because of the infinite dataset, we can only rely on an estimate of the full gradient.
\citet{harikandeh2015stopwasting} analysed this scenario under the assumptions of $z$ being concave, showing that SVRG is robust to an inexact computation of the full gradient. In particular, it is still possible to recover the original convergence rate if the error decreases at an appropriate rate. In Section \ref{sec:conv} we will show how the estimation accuracy impacts on the convergence results with a non-concave objective.
Finally, the non-stationarity of the optimization problem introduces a bias into the SVRG estimator in Eq.~\eqref{E:svrg.gradient.correction}.
To overcome this limitation we employ importance weighting~\citep[\eg][]{rubinstein1981simulation,precup2000eligibility} to correct the distribution shift.
%Finally, we employ importance weighting~\citep[\eg]{rubinstein1981simulation,precup2000eligibility} to guarantee proper sampling of trajectories.

@@

but the snapshot gradient is not exact and the gradient estimate used between snapshots is corrected using importance weighting:\footnote{\color{red} Note that $g$ can be any unbiased estimator w/o baseline. The unbiasedness is required for theoretical results (\eg Appendix~\ref{A:gradient_estimators}).}

@@

There are been a surge of work investigating variance reduction techniques for policy gradient methods.
% The most widespread technique of reducing the variance preserving the unbiasedness of the gradient is through baselines~\citep[\eg][]{williams1992simple,weaver2001optimal,peters2008reinforcement,Thomas2017actionbaseline,wu2018variance}.
The specific structure of the policy gradient allows to incorporated a baseline (\ie a function $b :\mathcal{S}\times\mathcal{A} \to \realspace$) without affect the unbiasedness of the gradient~\citep[\eg][]{williams1992simple,weaver2001optimal,peters2008reinforcement,Thomas2017actionbaseline,wu2018variance}.
Although the baseline can be arbitrarily selected, literature often refer to the optimal baseline as the one minimizing the variance of the estimate.
Nevertheless, even the baseline needs to be estimated from data. This fact may partially reduce its effectiveness by introducing variance.
Despite these approaches and SVRG has the same goal, they are substantially different.
In particular, the proposed SVRPG does not make explicit use of the structure of the policy gradient framework and it is independent from the underline gradient estimate (\ie w/o baseline).
This suggests that would be possible to integrate an ad-hoc SVRPG baseline in order to further reduce the variance of the estimate.
But since this paper is about the applicability of SVRG technique to RL we consider this topic as future work.
Additionally, the experiments show the advantage of using simple SVRPG against G(PO)MPD with baseline (see the \textcolor{blue}{half-cheetah} domain in Section~\ref{sec:exp}).

@@

For G(PO)MDP, we do not have a compact expression for $g$, but since it is derived from REINFORCE by neglecting some terms of the kind $\nabla \log \pi_{\vtheta}(a_t | s_t)\gamma^t r_t$ \cite{baxter2001infinite,peters2008reinforcement}, the above bound still holds.
